---
output: html_document
---

<a class="anchor" id="Index"></a>

```{r, echo=F, eval=T, message=F, warning=F, error=F, comment=F}
library('readxl')
library(ggplot2)
library(plotly)
library(forecast)
library(fontawesome)
source('~/API/plotly.R')

Sys.setenv("plotly_username"=username)
Sys.setenv("plotly_api_key"=plotly_api)
```

  
<div class="fluid-row" id="header" style="padding-top: 10%;">
  <p>
    <h4 style="text-align: center;">
This site is still under construction. Please continue to come back for frequent updates and improvements.
    </h4>  
    <h1 style="text-align: center;">`r fa('tools')`</h1>
  </p>
  <h1> About </h1> 
  <br>
  <p>
Hello World, I'm <b>Juliann</b>! I recently graduated from CUNY School of Professional Studies with a Master of Science degree in Data Science.  I am a skilled analyst with 5 cumulative years of data-driven experience and possess a strong background in acquiring, organizing, analyzing, interpreting and visualizing data. 
  </p>
  <p>
I am actively pursuing exciting, challenging, and meaningful job opportunities to put my <b>data science</b> and <b>machine learning</b> skills to work. When I am not programming, you can find me exploring the streets of NYC, enjoying a hot cup of coffee, or puzzling with friends.  
  </p>

<br>
<br>

<a class="anchor" id="Portfolio"></a>

<h1>Portfolio</h1>
  <br>
  <p>
Academic and personal data science projects and tutorials:  
  </p>
  <br>

<table width='100%'>
  <tr><h3>Time Series</h3></tr>
  <tr>
    <td>
      <h4><a href="time_series.html">Forecasting</a></h4>
      Forecasting ATM Cash Withdrawals & Plotly Tutorial. 
    </td>
    <td width='75%'>
    

    
```{r, echo=F, eval=T, message=F, warning=F, error=F, comment=F, fig.height=3, fig.width=4, out.width='100%', fig.align='center', cache=T}

load('./projects/load/TS.RData')

data <- data.frame(cbind(Day = seq(1, 365), Cash = as.vector(ATM1_ts)))
forecast <- data.frame(cbind(Day = seq(366, 397), Cash = as.vector(ATM1_fc$mean)))
level <- data.frame(cbind(Day = seq(366, 397), 
                          Upper = as.vector(ATM1_fc$upper[,2]), 
                          Lower = as.vector(ATM1_fc$lower[,2])))


p1 <- ggplot() +
  geom_line(data=data, aes(Day,Cash), size=.25)+
  geom_line(data=forecast, aes(Day,Cash), color="#000000", size=.75, alpha=.9)+
  geom_ribbon(data=level, aes(x=Day, ymax=Upper, ymin=Lower), fill='#C1B176', alpha=.4)+
  labs(title = "31-Day ATM Withdrawal Forecast", x="Days", y="Cash", caption="95% Confidence Interval")+
  theme_bw()+
  theme()
  
ggplotly(p1) %>% 
  layout(annotations = list(x = 1, y = -0.2, text = "95% Confidence Level",
                            showarrow = F, xref='paper', yref='paper', 
                            xanchor='right', yanchor='bottom', xshift=0, yshift=0,
                            font=list(size=10, color="grey")))
```

  </td>
  </tr>
</table>
<p>
<b>In the works:</b> Natural Language Processing, Recommender Systems, Spatial Analytics, Predictive Analytics, and more. 
</p>
</div>


