---
output: 
  html_document:
    highlight: monochrome 
    df_print: paged
    code_folding: "hide"
---

<div class="row align-items-center" style="padding-top: 10%;">

<h1 style="display: block; font-size: 35px; font-family: 'Courier New', 'mono'; text-align: left; margin: 0; line-height: 35px;" id="ABOUT"> Learn<b>R</b>: Time-Series Tutorial</h1>
</div>

<br>
<br>

### Getting Started
 
<p>
This step-by-step tutorial is modified from an academic <a href="https://github.com/jemceach/predictive-analytics/tree/master/Project_One" target="_blank" style="display: inline;"> project</a> designed for students to develop a one-month forecast for cash withdrawals from several ATM machines. For replication purposes, data can be <a href="https://github.com/jemceach/jemceach.github.io/blob/master/projects/data/ATM_Data.xlsx" target="_blank" style="display: inline;"> downloaded </a> from my portfolio repository. 
</p>
<br>

#### Loading Packages 

<p>
To begin, the following packages should be installed and loaded in R:
</p>

```{r libraries, echo = T, message=F, warning=F, tidy=T, error=F, class.source='fold-show'}
# Data preparation 
library(readxl); library(dplyr); library(tidyr); library(tibble) 

# Data visualization
library(ggplot2); library(ggfortify); library(grid); library(gridExtra); library(extrafont); library(default); library(plotly)

# Timeseries / Forecasting
library(stats); library(imputeTS); library(forecast)
```

```{r, echo=F}
# Options for R code chunks
knitr::opts_chunk$set(
  echo = T, 
  message=F, 
  warning=F, 
  error=F, 
  comment=NA, 
  size = "small", 
  fig.width=10, 
  fig.height = 3, 
  out.width='100%', 
  fig.align='left'
  )

# Add font options for base R plots & ggplots 
windowsFonts(FGB = windowsFont("Franklin Gothic Book"))
extrafont::loadfonts(quiet = T)

# Set default augments for ggplot2 `theme()`
default::default(theme_gray) <- list(base_family = "Franklin Gothic Book")

default::default(theme) <- list(
  axis.text.x = element_text(size=8, angle = 0, hjust = NULL),
  axis.text.y = element_text(size=8, angle = 90, hjust = 1),
  plot.title = element_text(color="#000000", size=10, face="bold"),
  legend.title = element_text(size=8, color="#868b8c", face="bold"),
  strip.text.x = element_text(size = 8, face="bold"),
  panel.spacing = unit(2, "lines")
  )

# Quick Colors
lite_purple <- "#a1a5e0"; 
lite_blue <- "#b5c1d0"; 
dark_purple <- "#8f84a7"
pink <- "#b186a2"
```

#### Data Preparation

The `atm_data` is stored as a long dataframe with 3 columns: `DATE`, `ATM`, and `Cash` (in hundreds of dollars). Our task is to come up with several, one-month forecasts, so let's first prepare the data by separating the ATM machines into a wide format. Then, we can inspect the structure of the new dataframe. 


```{r prep, class.source='fold-show'}
# Load Data
atm_data <- readxl::read_excel("./projects/data/ATM_Data.xlsx") 

# Clean dataframe
atm <- atm_data %>% 
  ## Create wide dataframe
  tidyr::spread(ATM, Cash) %>%
  ## Arrange date in ascending order
  dplyr::arrange(DATE) 
```

Our data is now structured into 4 columns and contains one year (365 observation) of dates and transactions.

```{r str, class.source='fold-show'}
str(atm)
```

<br>

### Data Exploration 

Before modeling the data, we need to understand the information we're working with. We can do so with summary statistic and graphic visualizations of the ATM data. 

<br>

#### Summary Statistics

Using the `summary` function, we can quickly confirm that our `DATE` variable ranges one year, from 01 May 2009 to 30 April 2010. ATM2 contains one missing observation and the max value on ATM3, compared to the other quartiles, suggests a potential outlier in the data. 

```{r, class.source='fold-show'}
summary(atm)
```

#### Visualization 

<p>Graphs are useful tools in the exploration process. Histograms allow us to visualize distrubtion of `ATM` cash withdrawals and scatterplots enable us to better understand the relationship between `DATE` and `ATM`.</p> <br> 

<p>The <em>histogram</em> plots shows that all ATM series contain are skewed left. The distribution of ATM1 and ATM3 appear bimodal, whereas ATM2 shows multimodal peaks. There is also a clear outlier in ATM3 that we will impute later on. </p>

```{r echo=F}
p_hist <- atm %>% 
  ## Gather observations for facet plot
  tidyr::gather(key=ATM, value=Cash, ATM1,ATM2, ATM3) %>%
  ## Plot atms as histogram with ggplot2 package
  ggplot(aes(Cash)) +
    # Create scatterplots
    geom_histogram(color=dark_purple, fill=lite_purple, alpha=.8) +
    # Wrap panels based on ATM
    facet_wrap(~ATM, scales='free', nrow=1) +
    # Modify labels
    labs(title="ATM Histogram",y="", x="")+
    scale_x_continuous(labels = scales::dollar_format())+
    # Call theme components
    theme_gray()+
    theme()

plotly::ggplotly(p_hist) 
```

<p>The <em>scatterplots</em> show the relationship between cash and date for each machine. ATM1 and ATM3 show similar non-linear trends, suggestive of a potential trend within the timeseries. The transaction on Feb 9 2010 for ATM3 stands out as a clear outlier to the typical pattern exhibited by the machine. ATM2 appears to follow a negative, linear trend until Feb 2010, where the withdrawals begin to increase. </p>
 
```{r echo=F}
# plot atms as scatterplot
p_scatter <- atm %>% 
  ## Gather observations for facet plot
  tidyr::gather(key=ATM, value=Cash, ATM1,ATM2, ATM3) %>%
  ## Plot with ggplot2 package 
  ggplot(aes(DATE, Cash)) +
    ## Create scatterplots
    geom_jitter(size=1, color=lite_purple, alpha=.8) +
    ## Plot regression line to see patterns
    geom_smooth(method="loess", size=1.25, color=dark_purple) +
    ## Wrap panels based on ATM
    facet_wrap(~ATM, scales='free', nrow=1) +
    ## Modify labels
    labs(title="ATM Scatterplot",x="", y="")+
    scale_y_continuous(labels = scales::dollar_format()) +
    ## Call theme components
    theme_gray()+
    theme()

ggplotly(p_scatter)
```


### Time Series Analysis

We will next use the `stats` package to create a time series from our data and examine its characteristics. 

<br>

#### Creating Time Series Objects

Using the `ts` function, we can coerce the ATM dataframe into a time series at a weekly frequency. This frequency makes the most sense as we have one year of data and our goal is to forecast one month of transactions. The time series plot below show high weekly variance as suspected in from the scatterplots above. 

```{r class.source='fold-show', cache=F, fig.height=4}
# Set outlier in ATM4 to NA for imputation 
atm$ATM3[which.max(atm$ATM3)] <- NA

# Create TS object with weekly frequency (with NA values)   
atm_ts_withNA <- dplyr::select(atm, -DATE) %>% 
  stats::ts(start=c(1), frequency = 7, class="ts")

# impute missing values with series mean 
atm_ts_imp <- atm_ts_withNA %>% imputeTS::na_mean()
```


#### TS Imputations

<br>

<p>In the code above, the missing value in ATM2 and extreme outlier in ATM3 were both substituted with the mean series value.  We can visualize the imputation process using `plotNA.imputations` function from the <a href="https://cran.r-project.org/web/packages/imputeTS/imputeTS.pdf" target=_blank" style=  "display: inline;">`imputeTS`</a> package. The plot below uses this function along with some custom formatting to hone in on the two changes made to the ATM series. This package also contains additional functions to view the gapsize and distribution of missing values. These features can be helpful for timeseries with many missing values.</p>

```{r}
# Formatting Function: Fill rectange to match ggplot dark theme 
plt_gray_theme=function() do.call(rect,as.list(c(par()$usr[c(1,3,2,4)],col="#ededed")))

# Initiate layout for NA plots
graphics::layout(mat = matrix(c(1, 2), nrow = 1, ncol = 2), heights = c(1, 1), widths = c(1, 1))

# Set Graphical Parameters
par(family="FGB", col="#000000", lwd=2.5, cex=.75, oma=c(0,0,2,0))

#Visualize Imputed Values
imputeTS::plotNA.imputations(
  atm_ts_withNA[,2], atm_ts_imp[,2], 
  xlim = c(20,30), ylim=c(0,180),
  main = "ATM2", xlab="Week", ylab="Cash",
  panel.first=plt_gray_theme(),
  colWithNA="#a098dc",
  colWithImputations="#c85b64",
  legend=F
  )

# Set Graphical Parameters
par(new=T, family="FGB", col="#000000", lwd=2.5, cex=.75, oma=c(0,0,2,0))

#Visualize Imputed Values
imputeTS::plotNA.imputations(
  atm_ts_withNA[,3], atm_ts_imp[,3], 
  xlim = c(36,46), ylim=c(0,180),
  main = "ATM3", xlab="Week", ylab="Cash",
  panel.first=plt_gray_theme(),
  colWithNA="#a098dc",
  colWithImputations="#c85b64",
  legend=F
  )

# Add title to plot margins
mtext("Visualization of Imputed Values", outer = T, cex = 1, font=2, family="FGB")
```

#### Autoplot

We can quickly plot our multivariate timeseries using the `autoplot` function. The time series plots show high weekly variance as suspected in from our scatterplots. 

```{r, eval=F}
# Autoplot MTS with facet wrap
p_weekly <- ggplot2::autoplot(atm_ts_imp, facet=T, size=.5, alpha=.5, color="#000000") +
  labs(title = "Daily ATM Transactions by Week", x="", y="")+
  scale_x_continuous(breaks = seq(1,52,by=4),
                     labels = seq(1,52, by=4))+
  theme_gray()+ 
  theme()

ggplotly(p_weekly)
```

```{r, echo=F, fig.height=4}
# Update theme default for TS plots
default(theme) <- list(
  panel.spacing = unit(1, "lines"), 
  strip.placement = "outside",
  axis.title = element_blank(), 
  axis.text.y = element_blank(),
  axis.ticks = element_blank())

# forecast package affects autoplot layout; 
# RDS of plot before loading forecast package

rds<-readRDS("./projects/load/p_weekly.RDS")
names(rds$data) <- c("Index", "Series", "Cash")
ggplotly(rds)
```

#### Seasonality

<p>The `subseries` plot in the `forecast` package allows us to look at seasonal patterns in mini-plots. The forecast package does not allow for multivariate analysis, but with a little manipulation we can bind the three series together to create a faceted view of the three ATMs. </p>

<br>

<p>The horizontal lines in each plot represent the weekly mean value for ATM transactions. The plots show that some days have much more seasonal variation over time than others.</p>

```{r, fig.height=4}
# Subseries plots
ss_plot1 <- forecast::ggsubseriesplot(x=atm_ts_imp[,1])
ss_plot2 <- forecast::ggsubseriesplot(x=atm_ts_imp[,2])
ss_plot3 <- forecast::ggsubseriesplot(x=atm_ts_imp[,3])

# Labeller: lookup table
wk_day <- c(
  `1.5` = "Sun", `2.5` = "Mon", `3.5` = "Tues", `4.5`= "Wed",  `5.5` = "Thurs", `6.5` = "Fri", `7.5` = "Sat"
  )
  
# Bind SS plot data to create manual facet wrap of multivariate series
ss_plots <- rbind(cbind(ss_plot1$data, series="ATM1"),
      cbind(ss_plot2$data, series="ATM2"), 
      cbind(ss_plot3$data, series="ATM3")) %>%
  ggplot(aes(time,y, group = season))+
  geom_line(size=.5, alpha=.5, color="#000000")+
  geom_line(ggplot2::aes_(y = ~avg), col = dark_purple, alpha=.75)+
  geom_vline(xintercept = 2:7, size=.25, color="grey50")+ 
  coord_cartesian(xlim = c(1.25,7.75))+
  scale_y_continuous(labels = scales::dollar_format())+
  facet_wrap(~series, nrow=3, scales = "free_y", strip.position = "right")+
  scale_x_continuous(breaks = c(1.5:7.5), labels = wk_day)+
  theme_gray()+
  theme()+
  labs(title="Subseries Plot")

ggplotly(ss_plots)
```


#### Autocorrelation 

<p>We can use ACF plots to vizualize autocorrelation within each time series. All three ACF plots show a large, positive lag beginning at `r7`, which slowly decreases over a multiple of 7. These lags are indicative of a strong, weekly trend, especially for ATM1 and ATM3. </p>

<br>
<p>The `ggACF` function graphs one series at a time. However, we can bind our series to create a facet view of all series at once. 
</p>

```{r}
# Generate ACF data
ACF1 <- forecast::ggAcf(atm_ts_imp[,1], size=1, lag=52)
ACF2 <- forecast::ggAcf(atm_ts_imp[,2], size=1, lag=52)
ACF3 <- forecast::ggAcf(atm_ts_imp[,3], size=1, lag=52)

# Bind ACF data
ACF_all <- rbind(
  cbind(ACF1$data, series="ATM1"), 
  cbind(ACF2$data, series="ATM2"), 
  cbind(ACF3$data, series="ATM3")
  ) 

# Table view: First 7 lags
ACF_all %>%
  select(lag, Freq, series) %>% 
  as.data.frame() %>% 
  filter(lag < 8) %>%
  mutate(series = as.character(series),
         Freq = round(Freq, 2)) %>% 
  spread(key=lag, value=Freq) %>%
  rename_if(is.numeric, list(~ paste0("r", .)))
```

<br>

<p>The plots also show a large, negative lag pattern, which begins at lag 2 and lag 5, and also decrease over a multiple of 7. The negative lags are strongest in the ATM2 series. These lags represent troughs which suggest the ATM transactions tend to be five and two days behind transaction peaks. </p>
<br>

<p>The autocorrelation in the series also indicates that the time-series contains non-stationary data. This can affect results of regression as the series do not fluxuate around a constant mean or with constant variance. We will apply two different methods in the next section to properly evaluate the data. </p>

```{r}
# Confidence interval lines
ci <- qnorm((1 + 0.95) / 2) / sqrt(nrow(ACF_all))

# Facet plot ACF
p_ACF_all <- ACF_all %>%
  ggplot(aes(lag, Freq))+
  geom_hline(yintercept = 0, size=1, alpha=.8)+
  geom_hline(yintercept = c(-ci, ci), color = lite_blue, linetype = "dashed", size=1, alpha=.8)+
  geom_segment(aes(x = lag, xend = lag, y = 0, yend = Freq), size=1, alpha=.8)+
  facet_wrap(~series, nrow=1)+
  theme_gray()+
  theme(axis.title.y = element_blank())+labs(x="\n\nlag")

ggplotly(p_ACF_all)
```

<br>

### Forecasting

We will compare two different forecasting approaches - STL and ARIMA. Usings a custom function, we can optimize gathering and plotting the multivariate series. 

```{r}
gather_fc<- function(series1, series2, series3){
  rbind(
    cbind(as.data.frame(series1), series="ATM1"), 
    cbind(as.data.frame(series2), series="ATM2"), 
    cbind(as.data.frame(series3), series="ATM3")
  )  %>% 
  as.data.frame() %>%
  mutate(series = as.character(series))%>%
  rowid_to_column("Index") %>%
  group_by(series)%>%
  mutate(Index = seq(as.Date("2010/5/1"), as.Date("2010/5/31"), "days"))
}

plt_fc <- function(gather_fc, title){
  TS_all <- atm_ts_imp %>%
    as.data.frame() %>% 
    rowid_to_column("Index") %>%
    dplyr::mutate(Index = seq(as.Date("2009/5/1"), as.Date("2010/4/30"), "days")) %>%
    tidyr::gather(key="series", value="value", -Index)
  ggplot() + 
  geom_line(data=TS_all, aes(Index, value))+ 
  geom_ribbon(data=gather_fc, 
              aes(x=Index, ymin=`Lo 95`, ymax=`Hi 95`), fill=lite_purple, alpha=.75)+
  geom_ribbon(data=gather_fc, 
              aes(x=Index, ymin=`Lo 80`, ymax=`Hi 80`), fill=dark_purple, alpha=.5)+
  geom_line(data=gather_fc, 
            aes(Index, `Point Forecast`), color=dark_purple, alpha=1)+ 
  scale_y_continuous(labels = scales::dollar_format())+
  labs(title=title)+
  facet_wrap(~series, nrow=3, scales = "fixed")+
  theme_gray()+
  theme(axis.title = element_blank(), 
        panel.spacing = unit(1, "lines"), strip.placement = "outside") 
}
```

<br>

```{r, echo=F}
#STL
STLF1 <- atm_ts_imp[,1] %>% stlf(h=31) 
STLF2 <- atm_ts_imp[,2] %>% stlf(h=31) 
STLF3 <- atm_ts_imp[,3] %>% stlf(h=31) 

#ARIMA
ARIMA1 <- auto.arima(atm_ts_imp[,1], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)
ARIMA2 <- auto.arima(atm_ts_imp[,2], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)
ARIMA3 <- auto.arima(atm_ts_imp[,3], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)
```

#### STL decomposition {.tabset}

STL stands for Seasonal and Trend decomposition using Loess. The `stlf` function decomposes a non-stationary series and generate a seasonally adjusted forecast. 

<br>

```{r fig.height=4}
STLF_all <- gather_fc(STLF1, STLF2, STLF3)
p_STLF <- plt_fc(STLF_all, title = "STLF Forecast")
ggplotly(p_STLF)
```


##### ATM1

```{r}
STLF1 <- atm_ts_imp[,1] %>% stlf(h=31)  
STLF1$model
```

##### ATM2

```{r}
STLF2 <- atm_ts_imp[,2] %>% stlf(h=31) 
STLF2$model
```

##### ATM3

```{r}
STLF3 <- atm_ts_imp[,3] %>% stlf(h=31) 
STLF3$model
```

#### ARIMA {.tabset}

ARIMA is an acronym for auto regressive integrated moving average. We can use `auto.arima()` to automatically select models based on the best information criterion. The augment `D` represents the order of seasonal-differencing. For the ATM models, we can `D=1` to account for seasonal differencing of our data to find the best fit for the ARIMA models.

```{r fig.height=4}
ARIMA_all <- gather_fc(ARIMA1, ARIMA2, ARIMA3)
p_ARIMA <- plt_fc(ARIMA_all, title = "ARIMA Forecast")
ggplotly(p_ARIMA)
```

##### ATM1

```{r}
ARIMA1 <- auto.arima(atm_ts_imp[,1], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)

ARIMA1$model
```

##### ATM2

```{r}
ARIMA2 <- auto.arima(atm_ts_imp[,2], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)

ARIMA2$model
```

##### ATM3

```{r}
ARIMA3 <- auto.arima(atm_ts_imp[,3], D = 1, lambda = "auto", approximation = F, stepwise = T) %>%
  forecast(h=31)

ARIMA3$model
```




